A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2024-10-10 manuscript.qmd
go over mixture model formalism (WIP)

# Introduction
Head and Neck Squamous Cell Carcinoma (HNSCC) are known to spread through the lymphatic system  often leading to metastases in the lymph nodes [@mukherji_cervical_2001;@shah_patterns_1990].  To minimize nodal recurrences, lymph node levels (LNL) at risk of harboring occult metastases are typically irradiated electively.  Current guidelines for different tumor locations are based on the overall prevalence of nodal disease as reported in literature [@biau_selection_2019;@mukherji_cervical_2001;@shah_patterns_1990].

To personalize the prediction of the risk of occult metastases, given a patient's individual diagnosis, we previously published a large, multi-centric dataset where the lymphatic involvement per LNL is available for each patient[@ludwig_dataset_2022;@ludwig_multi-centric_2023].Building on this dataset, we introduced an interpretable hidden Markov model (HMM), trained to predict the risk for occult nodal disease, given an individual patient's diagnosis [@ludwig_hidden_2021].

@@

Initially, separate models were trained for distinct tumor locations, such as the oropharynx and oral cavity. These tumor locations are also used in guidelines to define the elective target volumes [@biau_selection_2019]. However, this approach did not account for variations in lymphatic spread between subsites within these tumor regions. With data for more than 2700 patients available, we can now further analyze subsite specific spread patterns. Closer analysis showed that pooling subsites into a single model led to inaccurate predictions, as it failed to capture distinct lymphatic spread patterns. To resolve this, we propose using a mixture of HMMs, which allows us to model the lymphatic spread more accurately for tumors located near anatomical borders, such as those between the oropharynx and oral cavity (e.g., tumors in the palate).

Additionally, we extend the analysis to a broader mixture model that encompasses tumors of the oral cavity, oropharynx, hypopharynx, and larynx, offering a more comprehensive understanding of lymphatic spread across these regions.

@@

The datasets 1-4 are publicly available as CSV tables (@ludwig_multi-centric_2023 @ludwig_detailed_2022) and can be interactively explored on [LyProX](https://lyprox.org). For each patient the primary tumor subsite is reportd and each indicidual LNL is reported as either metastatic or healthy given the available diagnostic modalities, which include pathology after neck dissection in some patients. 
In this work we will stratify the tumor locations into different ICD codes which are depicted in @fig-subsites.

@@

The HMM describes each LNL $v \in 1, 2, ..., V$ by a binary random variable corresponding to the status of the LNL; healthy (0) or involved (1). The entire state of a patient with $V$ LNLs is defined by the $v$-dimensional vector $\mathbf{X} = [X_1, X_2, ...,X_V]$. In the HMM, a patient's involvement is modeled over time $t$. Thus, a patient's state of lymph node involvement $\mathbf{X}[t]$ evolves over discrete time steps $t$. Let us enumerate all $2^V$ possible states, representing all combinations of LNL involvement. In this paper, we consider ipsilateral LNLs I, II, III, IV and V, which amounts to 32 possible states. The HMM is then specified by a transition matrix $\mathbf{A}$:

@@

The mixture model assumes that the data is generated by a set of $M$ different lymphatic spread models. Each patient $k$ is then assumed to be generated by one specific model from this set of $M$ models.

To formalize this, each patient $k$ can be assigned to a specific model via a binary latent vector $\boldsymbol{\epsilon}_k$. The vector $\boldsymbol{\epsilon}_k$ has length $M$, with exactly one element set to 1, indicating which model generated the patient’s data, and all other elements set to 0. Thus, for patient $k$, the latent variable $\boldsymbol{\epsilon}_k$ can be interpreted as a categorical indicator variable that encodes the assignment to one of the $M$ lymphatic spread models.

Next, we introduce subsite-specific mixing parameters $\boldsymbol{\pi}^s = \{\pi_1^s, \pi_2^s, \dots, \pi_M^s\}$, where each $\pi_m^s$ represents the prior probability that a patient with tumors in subsite $s$ is generated by model $m$. These mixing parameters must satisfy the condition:

@@

The mixing parameters $\boldsymbol{\pi}^s$ determine the proportion of patients with tumors in subsite $s$ that are expected to be generated by each model. Specifically, $\pi_m^s$ is the probability that a randomly chosen patient with a tumor in subsite $s$ is generated by model $m$.

The latent variables $\boldsymbol{\epsilon}_{k}$, on the other hand, are patient-specific and represent the assignment of each patient to a particular model. The latent variables are central to the model, as they are hidden (unobserved) and must be inferred.

>>>>

# Introduction

Head and neck squamous cell carcinomas (HNSCCs) are known to spread through the lymphatic system often leading to metastases in the lymph nodes [@mukherji_cervical_2001;@shah_patterns_1990]. To minimize nodal recurrences, lymph node levels (LNLs) at risk of harboring occult metastases are typically irradiated electively. Current guidelines for different tumor locations are based on the overall prevalence of nodal disease as reported in literature [@biau_selection_2019;@mukherji_cervical_2001;@shah_patterns_1990].

To personalize the prediction of the risk of occult metastases, given a patient's individual diagnosis, we previously published a large, multi-centric dataset where the lymphatic involvement per LNL is available for each patient[@ludwig_dataset_2022;@ludwig_multi-centric_2023]. Building on this dataset, we introduced an interpretable hidden Markov model (HMM), trained to predict the risk for occult nodal disease, given an individual patient's diagnosis [@ludwig_hidden_2021].

@@

Initially, separate models were trained for distinct tumor locations, such as the oropharynx and oral cavity. These tumor locations are also used in guidelines to define the elective target volumes [@biau_selection_2019]. However, this approach did not account for variations in lymphatic spread between subsites within these tumor regions. With data from more than 2700 patients available, we can now further analyze subsite specific spread patterns. Closer analysis showed that pooling subsites into a single model led to inaccurate predictions, as it failed to capture distinct lymphatic spread patterns. To resolve this, we propose using a mixture of HMMs, which allows us to model the lymphatic spread more accurately for tumors located near anatomical borders, such as those between the oropharynx and oral cavity (e.g., tumors in the palate).

Additionally, we extend the analysis to a broader mixture model that encompasses tumors of the oral cavity, oropharynx, hypopharynx, and larynx, resulting in further personalized predictions of lymphatic spread across these regions.

@@

The datasets 1-4 are publicly available as CSV tables [@ludwig_multi-centric_2023;@ludwig_detailed_2022] and can be interactively explored on [LyProX](https://lyprox.org). For each patient the primary tumor subsite is reportd and each indicidual LNL is reported as either metastatic or healthy given the available diagnostic modalities, which include pathology after neck dissection in some patients. 
In this work we will stratify the tumor locations into different ICD codes which are depicted in @fig-subsites.

@@

The HMM describes each LNL $v \in 1, 2, ..., V$ by a binary random variable corresponding to the status of the LNL; healthy (0) or involved (1). The entire state of a patient with $V$ LNLs is defined by the $V$-dimensional vector $\mathbf{X} = [X_1, X_2, ...,X_V]$. In the HMM, a patient's involvement is modeled over time $t$. Thus, a patient's state of lymph node involvement $\mathbf{X}[t]$ evolves over discrete time steps $t$. Let us enumerate all $2^V$ possible states, representing all combinations of LNL involvement. In this paper, we consider ipsilateral LNLs I, II, III, IV and V, which amounts to 32 possible states. The HMM is then specified by a transition matrix $\mathbf{A}$:

@@

The mixture model assumes that the data is generated by a set of $M$ different lymphatic spread models. Each patient $k$, with their primary tumor in subsite $s \in (1, 2, \ldots, S)$, is assumed to be generated by one specific model $m \in (1, 2, \ldots, M)$ from this set of $M$ models with probability $\pi_m^s$. These so-called _mixing parameters_ $\boldsymbol{\pi}^s = \{\pi_1^s, \pi_2^s, \dots, \pi_M^s\}$ must satisfy the condition

@@

If we could record the componet $m$ from which a patient $k$ was drawn from, we could store this information in a binary latent vector $\boldsymbol{\epsilon}_k$. The vector $\boldsymbol{\epsilon}_k$ has length $M$, with exactly the $m$-th element set to 1, indicating which model generated the patient’s data, and all other elements set to 0. Thus, for patient $k$, the latent variable $\boldsymbol{\epsilon}_k$ can be interpreted as a categorical indicator variable that encodes the assignment to one of the $M$ lymphatic spread models. Unfortunately, the latent variables are hidden (unobserved) and must be inferred.
